Comparison of Fast, Hybrid Imaging Architectures for Multi-scale, Hierarchical Aperture Arrays

### 2024-10-31 ms.tex
Remove boldfacing

\begin{abstract}
Two major areas of modern radio astronomy, namely, explosive astrophysical transient phenomena and observations of cosmological structures, are driving \textbf{the design of} aperture arrays towards large \textbf{numbers of} low-cost elements consisting of multiple spatial scales \textbf{spanning the dimensions of individual elements, the size of stations (groupings of individual elements), and the spacing between stations}. Such multi-scale, hierarchical aperture arrays require a combination of data processing architectures-- pre-correlation beamformer, generic version of FFT-based direct imager, post-correlation beamformer, and post-correlation FFT imager-- operating on different \textbf{ranges of} spatial scales to obtain optimal performance in imaging the entire field of view. Adopting a computational cost metric based on the number of floating point operations, its distribution over the dimensions of discovery space, namely, field of view, angular resolution, polarisation, frequency, and time is examined to determine the most efficient hybrid architectures over the parameter space of hierarchical aperture array layouts. Nominal parameters of specific upcoming and planned arrays-- the SKA at low frequencies (\texttt{SKA-low}), \texttt{SKA-low-core}, a long baseline extension to \texttt{SKA-low} (\texttt{LAMBDA-I}), compact all-sky phased array (\texttt{CASPA}), and a lunar array (\texttt{FarView-core})-- are used to determine the most optimal architecture hierarchy for each from a computational standpoint, and provide a guide for designing hybrid architectures for multi-scale aperture arrays. For large, dense-packed layouts, a FFT-based direct imager is most efficient for most cadence intervals, and for other layouts that have relatively lesser number of elements or greater sparsity in distribution, the best architecture is more sensitive to the cadence interval, which in turn is determined by the science goals.
\end{abstract}

@@

\textbf{The variable and transient radio sky encompasses a vast range of astrophysical phenomena. Notable examples include pulsars, stellar bursts \citep{Zhang+2023}, fast radio bursts (FRBs), magnetars, rotating radio transients (RRATs), gamma-ray burst (GRB) afterglows, Jovian and solar bursts, flare stars, cataclysmic variables (CVs), exoplanet and stellar outbursts, X-ray binaries, novae, supernovae, active galactic nuclei (AGNs), blazars, tidal disruption events, and counterparts to gravitational wave (GW) events \citep{Lorimer+2007,Thyagarajan+2011,Keane2013,Thornton+2013,Bochenek+2020}. Their timescales span orders of magnitude, from nanoseconds to days, across wide frequency ranges and \textbf{polarisation states} \citep[][and references therein]{Pietka+2015,Chandra+2016,Nimmo+2022}. For example, pulsars and their giant radio pulses can operate on timescales from milliseconds to seconds, with millisecond pulsars on milliseconds to tens of milliseconds, while FRBs are active on sub-millisecond to tens of millisecond timescales \citep{Crawford+2022,Gupta+2022,Snelders+2023}. Extremely energetic and ultra-fast phenomena, such as sub-millisecond pulsars and magnetars \citep{Du+2009,Haskell+2018}, and nanosecond-scale giant pulses from the Crab pulsar \citep{Hankins+2003,Hankins+2007,Eilek+2016,Philippov+2019}, can reveal the fundamental nature of matter and exotic physics. The effectiveness of discovering these transients depends on the instrument's field of view and the duration of on-sky observation \citep{Cordes2007}, ideally requiring continuous real-time operation over a large instantaneous field of view to capture phenomena ranging from microseconds to days.}

% The variable and transient radio sky encompasses a variety of intriguing astrophysical phenomena with a range of timescales spanning orders of magnitude, wide frequency ranges, and \textbf{polarisation states} \citep[][and references therein]{Pietka+2015,Chandra+2016,Nimmo+2022}. Notable examples include pulsars and their giant radio pulses, magnetars, fast radio bursts (FRBs), rotating radio transients (RRATs), Jovian and solar bursts, flare stars, cataclysmic variables (CVs), exoplanet and stellar outbursts, X-ray binaries, novae, supernovae, gamma-ray bursts (GRBs) and their afterglows, active galactic nuclei (AGNs) and blazars, tidal disruption events, both the prompt and afterglow counterparts to gravitational wave (GW) merger events, as well as many phenomena yet to be understood \citep{Lorimer+2007,Thyagarajan+2011,Keane2013,Thornton+2013,Bochenek+2020}. The effectiveness of discovering these transients depends on the instrument's field of view and the duration of on-sky observation \citep{Cordes2007}. Ideally, instruments should operate continuously and in real time, covering a wide range of timescales from microseconds to days, with a large instantaneous field of view.

Another significant goal of modern radio astronomy is the exploration of structure formation and constraining of cosmological parameters in the early Universe using redshifted 21~cm line of neutral Hydrogen as a tracer of large-scale structures \citep[][and references therein]{Morales+2010,Pritchard+2012}. One prominent example includes probing the Cosmic Dawn and the Epoch of Reionisation signifying the appearance of the first luminous objects in the Universe and their impact on structure formation in the Universe at $z\gtrsim 6$. Another example is to probe the Dark Energy and acceleration of the Universe at $1\lesssim z\lesssim 4$ using intensity mapping of redshifted 21~cm from H{\sc i} bound to galaxies \citep{CosmicVisions+2018}. Such cosmological radio observations probing large-scale structures through spatial power spectrum or imaging require interferometer arrays of large fields of view \textbf{(hundreds to thousands of square degrees)}, dense sampling of the large-scale spatial modes \textbf{(tens of kpc to tens of Mpc)}, and high sensitivity \textbf{($\lesssim 1$~mK noise levels)}.

@@

The hierarchical, multi-scale aperture array scenario considered involves spatial scales corresponding to the dimensions of the collecting element, station (a grouping of elements), and the array (a grouping of stations), denoted by $D_\textrm{e}$, $D_\textrm{s}$, and $D_\textrm{A}$, respectively. \textbf{Figure~\ref{fig:geometry-notation} illustrates the geometry and notation for hierarchical multi-scale aperture arrays studied in this paper.}

\begin{figure}
\includegraphics[width=\linewidth]
  % {figures/aperture_image_dimensions.pdf}
{fig_01.pdf}
\caption{\textbf{Geometric view of a hierarchical multi-scale aperture array, which consists of $N_\textrm{s}$ stations, each a collection of $N_\textrm{eps}$ elements. Stations, $m$ and $n$, at locations, $\boldsymbol{r}_m$ and $\boldsymbol{r}_n$, respectively, relative to the origin, $\mathcal{O}$, are separated by $\Delta\boldsymbol{r}_{mn}$. Elements, $a$ and $b$, in stations, $m$ and $n$, are denoted by $a_m$ and $b_n$, located at $\boldsymbol{r}_{a_m}$ and $\boldsymbol{r}_{b_n}$, respectively. $\Delta\boldsymbol{r}_{a_m b_n}$ is the separation between these elements. The spatial sizes of the element, station, and the array in the aperture plane are denoted by $D_\textrm{e}$, $D_\textrm{s}$, and $D_\textrm{A}$, respectively. The solid angle, $\Omega_\textrm{e}\sim (\lambda/D_\textrm{e})^2$, corresponding to the element size determines the overall field of view in the sky plane accessible by the aperture array. Intra-station architectures produce station-sized virtual telescopes that simultaneously produce beams of solid angle, $\Omega_\textrm{s}\sim (\lambda/D_\textrm{s})^2$, in multiple directions, $\hat{\boldsymbol{s}}_k$, that can fill $\Omega_\textrm{e}$. Inter-station processing towards any of these beams produces virtual telescopes of size, $D_\textrm{A}$, that simultaneously produce beams of solid angle, $\Omega_\textrm{A}\sim (\lambda/D_\textrm{A})^2$, in multiple directions, $\boldsymbol{\sigma}_\ell$, relative to $\hat{\boldsymbol{s}}_k$, that can fill the solid angle of that beam.} 
\label{fig:geometry-notation}}
\end{figure}

The number of stations in the array and the number of elements per station (indexed by $m$) are denoted by $N_\textrm{s}$ and $N_{\textrm{eps}_m}$, respectively. The number of elements in all stations are assumed to be equal, $N_{\textrm{eps}_m} = N_\textrm{eps}$, for convenience. \textbf{Without loss of generality, this study can be extended to arrays with heterogeneous station densities and layouts such as LOFAR, but the idiosyncrasies of the array will have to be taken into account.}

The specific aperture array examples considered are the \texttt{SKA-low}, the core of the \texttt{SKA-low} (\texttt{SKA-low-core}), the first phase of \textbf{the concept of} a long baseline extension to \texttt{SKA-low} called Low-frequency Australian Megametre Baseline Demonstrator Array (\texttt{LAMBDA-I}), the Compact All-Sky Phased Array \citep[\texttt{CASPA}\footnote{Presently, CASPA's stations are intended for localisation and astrometry, rather than for aperture synthesis.};][]{Luo+2024}, and the core of a lunar array \citep[\texttt{FarView};][]{Polidan+2024} which is hereafter referred to as \texttt{FarView-core}. \texttt{LAMBDA-I}, \texttt{CASPA}, and \texttt{FarView-core} are hypothetical interferometers \textbf{that are still in concept phase.}

@@

\textbf{Note that, traditionally, cosmological applications such as probing the epoch of reionisation (EoR) using redshifted 21~cm have relied on storing the visibilities and processing them offline. In such cases, although imaging costs like gridding, Discrete Fourier Transform (DFT), or FFT will not apply, the correlations must still be obtained in real-time. With large arrays, the traditional correlator approach may be too expensive. For example, the Packed Ultra-wideband Mapping Array \citep[PUMA;][]{PUMA+2019}, a planned radio telescope for cosmology and transients with $\sim 5000-32000$ elements, is considering an FFT correlator architecture, a specific adaptation of the EPIC architecture for redundant arrays. This study, with a breakdown of costs for different operations, is also relevant for such cosmological arrays.}

@@

The primary metric for computational cost used in this study is the number of real floating point operations (FLOP) and the density of such operations in the discovery phase space volume whose dimensions span time, frequency, polarisation, and location on the sky. The computational cost will be estimated per independent interval of time (in FLOP per second or FLOPS) for all possible polarisation states per frequency channel ($\delta\nu$) per \textbf{independent angular resolution element} ($\delta\Omega$). In this paper, a discovery phase space volume element comprising of a single frequency channel at a single two-dimensional \textbf{angular resolution element (independent pixel)} will be referred to as a \textit{voxel} ($\delta\nu\,\delta\Omega$).

@@

Two broad classes of two-stage fast imaging architectures are considered. In both classes, the first stage is intra-station interferometry of the element electric fields using the options of voltage beamforming (BF), E-field Parallel Imaging Correlator (EPIC), beamforming of correlations (XBF), or correlation followed by FFT (XFFT). The second stage \textbf{involves} inter-station combination of the data products from the first stage. The main difference between the two classes is in whether the second stage uses the first stage products coherently or incoherently. \textbf{Although aperture arrays consisting of only three hierarchical scales are considered here, the analysis can be extended without loss of generality to arrays of even larger levels of spatial hierarchy.} A schematic of these hybrid architectures, their computational cost components, \textbf{and their extension methodology to multiple stages} are illustrated in Figure~\ref{fig:hybrid-architectures-cost-schematic}.

\begin{figure}
\includegraphics[width=0.95\linewidth]
  % {figures/flops_calc_abridged_v2.pdf}
{fig_02.pdf}
\caption{A hybrid, two-stage architecture for hierarchical aperture arrays \textbf{showing the methodology for extension to multiple stages.} Black arrows indicate pathways where the signal consists of electric fields with phase coherence. Gray arrows indicate pathways of a ``squared'' or a correlated signal with no absolute but only relative phase information. At the top are intra-station architectures, namely, BF, correlator, and EPIC. The top box denotes intra-station imaging options. The bottom box denotes inter-station imaging architectures that process coherent outputs obtained from multiple stations deploying intra-station BF and/or EPIC architectures. \textbf{The electric fields phase coherently synthesised by BF or EPIC at any stage can form the input for the next stage, thereby allowing the extension of the hierarchy.}
\label{fig:hybrid-architectures-cost-schematic}}
\end{figure}

@@

\textbf{In the discussions and equations that follow, a real-time calibration is assumed to have been performed on the input voltages. By assuming the system is reasonably stable, a much less frequent ($\ll 1/t_\textrm{acc}$) calibration will be sufficient, which therefore will not significantly affect the total cost budget \citep{Beardsley+2017,Gorthi+2021}. Although in principle an accurate calibration is possible in real time, applications such as EoR that require high-accuracy calibration have, for practical reasons, relied on offline calibration.}

@@

% The intra-station architectures comprising the first stage form the basis for both approaches.
This first stage essentially synthesises station-sized apertures with simultaneous, multiple virtual pointings. These station-sized synthesised apertures from the first stage will act as building blocks for the second stage of inter-station processing. The element, $a$, in station, $m$ is denoted by $a_m$. \textbf{The element and station locations are denoted by $\boldsymbol{r}_{a_m}$ and $\boldsymbol{r}_m$, respectively. Similarly, $\Delta\boldsymbol{r}_{a_m b_n}$ is the separation between an elements, $a$ and $b$, in stations, $m$ and $n$, respectively. $\Delta\boldsymbol{r}_{mn}$ denotes the station separation. This notation is illustrated in Figure~\ref{fig:geometry-notation}.} A summary of the intra-station costs are provided in Table~\ref{tab:intra-inter-station-coherent-imaging}.

@@

\textbf{Let $\widetilde{E}_{a_m}^{p}(\nu)$ represent the calibrated and potentially noise-weighted electric field at frequency, $\nu$, in polarisation state, $p$, measured by a station element, $a_m$, in station, $m$. Because each spectral channel is assumed to be processed independently, specifying the frequency can be dropped for convenience to write $\widetilde{E}_{a_m}^{p}(\nu)\equiv \widetilde{E}_{a_m}^{p}$. Then, such electric fields from individual station elements}
% The calibrated electric fields, $\widetilde{E}_{a_m}^p$, recorded by individual elements, $a_m$, within a station, $m$, and polarisation state, $p$,
can be phase-coherently superposed \textbf{to obtain the electric field in polarisation state, $\alpha$,} towards any desired direction as
\begin{align}
    \widetilde{\mathcal{E}}_m^\alpha(\hat{\boldsymbol{s}}_k) &= \sum_{a_m} \sum_p  \widetilde{\mathcal{W}}_{a_m}^{\alpha p *}(\hat{\boldsymbol{s}}_k) \, \widetilde{E}_{a_m}^p \, e^{i\frac{2\pi}{\lambda} \hat{\boldsymbol{s}}_k\cdot\boldsymbol{r}_{a_m}} \, , \label{eqn:intra-station-pol-hol-img-expl}
\end{align}
where, $\hat{\boldsymbol{s}}_k$ denotes the direction of the superposed beam, $k$,
% $\widetilde{E}_{a_m}^{p}$ represents the calibrated and potentially noise-weighted electric field measured by the station element, $a_m$,
and $\widetilde{\mathcal{W}}_{a_m}^{{\alpha p}^*}(\hat{\boldsymbol{s}}_k)$ denotes a complex-valued directional weighting\footnote{One can choose $\widetilde{\mathcal{W}}_{a_m}^{\alpha p}(\hat{\boldsymbol{s}}_k)=\mathcal{W}_{a_m}^{p\alpha}(\hat{\boldsymbol{s}}_k)$, the directional electric field sensitivity, if the signal-to-noise ratio is to be optimised. But it can also be generically chosen depending on the property desired of the estimator \cite[][]{Morales2011}.} \textbf{with the superscript indices $\alpha, p$, representing the contribution of polarisation state, $p$ in the station element, $a_m$, to polarisation state, $\alpha$, towards direction, $\hat{\boldsymbol{s}}_k$. The * denotes complex conjugate. Equation~(\ref{eqn:intra-station-pol-hol-img-expl}) resembles a DFT of the calibrated Electric field measurement after applying a complex-valued weighting.}

The polarised intensity in the \textbf{beamformed} pixel is then obtained by
\begin{align}
    \widetilde{\mathcal{I}}^{\alpha\beta}_m(\hat{\boldsymbol{s}}_k) &= \left\langle \widetilde{\mathcal{E}}_m^\alpha(\hat{\boldsymbol{s}}_k) \,  \widetilde{\mathcal{E}}_m^{\beta *}(\hat{\boldsymbol{s}}_k) \right\rangle \, , \label{eqn:intra-station-opt-pol-img-outprod}
\end{align}
where, angular brackets denote a temporal averaging across an interval of $t_\textrm{acc}$. \textbf{$\widetilde{\mathcal{I}}^{\alpha\beta}_m(\hat{\boldsymbol{s}}_k)$ results from the outer product of indices $\alpha$ and $\beta$. Superscript, $\alpha\beta$, thus denotes the four pairwise combinations of polarisation states of the intensity.} Even though it is an outer product over the polarisation indices, it will be referred to as a ``squaring'' operation, hereafter, for convenience because of what it reduces to if $\alpha=\beta$.

@@

Figure~\ref{fig:1D-incoherent-compcost-BF-LAMBDA-I} shows a breakdown of the computational cost per station per voxel for voltage beamforming as a function of the different station parameters. In each panel, all parameters except the one on the $x$-axis are kept fixed at the characteristic values of the \texttt{LAMBDA-I}, \textbf{\texttt{SKA-low}, and \texttt{SKA-low-core}} stations as highlighted in cyan. The total cost is dominated by the \textbf{DFT} beamforming in Equation~(\ref{eqn:intra-station-pol-hol-img-expl}) denoted by black dashed lines. \textbf{The DFT} beamforming cost per voxel scales linearly with the number of elements per station. The costs of squaring and temporal averaging in Equation~(\ref{eqn:intra-station-opt-pol-img-outprod}) are subdominant. Because the \textbf{squaring and temporal averaging} are performed on a per-pixel basis and not dependent on $N_\textrm{eps}$, they remain constant across $N_\textrm{eps}$. And because the number of independent pixels covering the field of view scales as $n_\textrm{bs} \simeq (D_\textrm{s}/D_\textrm{e})^2$, the cost per independent pixel remains constant across $D_\textrm{s}$ and $D_\textrm{e}$. All these operations have to be performed at every time step, $\delta t$, and hence remain constant regardless of $t_\textrm{acc}$. Although spatial averaging across multiple stations (orange dot-dashed line) is not a station-level operation, it is shown to emphasise that it constitutes only a negligible cost relative to the other components, noting that it needs to be only performed on a slower cadence of $t_\textrm{acc}$ on a per-pixel basis, and thus scales inversely with $t_\textrm{acc}$.

>>>>

\begin{abstract}
Two major areas of modern radio astronomy, namely, explosive astrophysical transient phenomena and observations of cosmological structures, are driving the design of aperture arrays towards large numbers of low-cost elements consisting of multiple spatial scales spanning the dimensions of individual elements, the size of stations (groupings of individual elements), and the spacing between stations. Such multi-scale, hierarchical aperture arrays require a combination of data processing architectures-- pre-correlation beamformer, generic version of FFT-based direct imager, post-correlation beamformer, and post-correlation FFT imager-- operating on different ranges of spatial scales to obtain optimal performance in imaging the entire field of view. Adopting a computational cost metric based on the number of floating point operations, its distribution over the dimensions of discovery space, namely, field of view, angular resolution, polarisation, frequency, and time is examined to determine the most efficient hybrid architectures over the parameter space of hierarchical aperture array layouts. Nominal parameters of specific upcoming and planned arrays-- the SKA at low frequencies (\texttt{SKA-low}), \texttt{SKA-low-core}, a long baseline extension to \texttt{SKA-low} (\texttt{LAMBDA-I}), compact all-sky phased array (\texttt{CASPA}), and a lunar array (\texttt{FarView-core})-- are used to determine the most optimal architecture hierarchy for each from a computational standpoint, and provide a guide for designing hybrid architectures for multi-scale aperture arrays. For large, dense-packed layouts, a FFT-based direct imager is most efficient for most cadence intervals, and for other layouts that have relatively lesser number of elements or greater sparsity in distribution, the best architecture is more sensitive to the cadence interval, which in turn is determined by the science goals.
\end{abstract}

@@

The variable and transient radio sky encompasses a vast range of astrophysical phenomena. Notable examples include pulsars, stellar bursts \citep{Zhang+2023}, fast radio bursts (FRBs), magnetars, rotating radio transients (RRATs), gamma-ray burst (GRB) afterglows, Jovian and solar bursts, flare stars, cataclysmic variables (CVs), exoplanet and stellar outbursts, X-ray binaries, novae, supernovae, active galactic nuclei (AGNs), blazars, tidal disruption events, and counterparts to gravitational wave (GW) events \citep{Lorimer+2007,Thyagarajan+2011,Keane2013,Thornton+2013,Bochenek+2020}. Their timescales span orders of magnitude, from nanoseconds to days, across wide frequency ranges and polarisation states \citep[][and references therein]{Pietka+2015,Chandra+2016,Nimmo+2022}. For example, pulsars and their giant radio pulses can operate on timescales from milliseconds to seconds, with millisecond pulsars on milliseconds to tens of milliseconds, while FRBs are active on sub-millisecond to tens of millisecond timescales \citep{Crawford+2022,Gupta+2022,Snelders+2023}. Extremely energetic and ultra-fast phenomena, such as sub-millisecond pulsars and magnetars \citep{Du+2009,Haskell+2018}, and nanosecond-scale giant pulses from the Crab pulsar \citep{Hankins+2003,Hankins+2007,Eilek+2016,Philippov+2019}, can reveal the fundamental nature of matter and exotic physics. The effectiveness of discovering these transients depends on the instrument's field of view and the duration of on-sky observation \citep{Cordes2007}, ideally requiring continuous real-time operation over a large instantaneous field of view to capture phenomena ranging from microseconds to days.

% The variable and transient radio sky encompasses a variety of intriguing astrophysical phenomena with a range of timescales spanning orders of magnitude, wide frequency ranges, and polarisation states \citep[][and references therein]{Pietka+2015,Chandra+2016,Nimmo+2022}. Notable examples include pulsars and their giant radio pulses, magnetars, fast radio bursts (FRBs), rotating radio transients (RRATs), Jovian and solar bursts, flare stars, cataclysmic variables (CVs), exoplanet and stellar outbursts, X-ray binaries, novae, supernovae, gamma-ray bursts (GRBs) and their afterglows, active galactic nuclei (AGNs) and blazars, tidal disruption events, both the prompt and afterglow counterparts to gravitational wave (GW) merger events, as well as many phenomena yet to be understood \citep{Lorimer+2007,Thyagarajan+2011,Keane2013,Thornton+2013,Bochenek+2020}. The effectiveness of discovering these transients depends on the instrument's field of view and the duration of on-sky observation \citep{Cordes2007}. Ideally, instruments should operate continuously and in real time, covering a wide range of timescales from microseconds to days, with a large instantaneous field of view.

Another significant goal of modern radio astronomy is the exploration of structure formation and constraining of cosmological parameters in the early Universe using redshifted 21~cm line of neutral Hydrogen as a tracer of large-scale structures \citep[][and references therein]{Morales+2010,Pritchard+2012}. One prominent example includes probing the Cosmic Dawn and the Epoch of Reionisation signifying the appearance of the first luminous objects in the Universe and their impact on structure formation in the Universe at $z\gtrsim 6$. Another example is to probe the Dark Energy and acceleration of the Universe at $1\lesssim z\lesssim 4$ using intensity mapping of redshifted 21~cm from H{\sc i} bound to galaxies \citep{CosmicVisions+2018}. Such cosmological radio observations probing large-scale structures through spatial power spectrum or imaging require interferometer arrays of large fields of view (hundreds to thousands of square degrees), dense sampling of the large-scale spatial modes (tens of kpc to tens of Mpc), and high sensitivity ($\lesssim 1$~mK noise levels).

@@

The hierarchical, multi-scale aperture array scenario considered involves spatial scales corresponding to the dimensions of the collecting element, station (a grouping of elements), and the array (a grouping of stations), denoted by $D_\textrm{e}$, $D_\textrm{s}$, and $D_\textrm{A}$, respectively. Figure~\ref{fig:geometry-notation} illustrates the geometry and notation for hierarchical multi-scale aperture arrays studied in this paper.

\begin{figure}
\includegraphics[width=\linewidth]
  % {figures/aperture_image_dimensions.pdf}
{fig_01.pdf}
\caption{Geometric view of a hierarchical multi-scale aperture array, which consists of $N_\textrm{s}$ stations, each a collection of $N_\textrm{eps}$ elements. Stations, $m$ and $n$, at locations, $\boldsymbol{r}_m$ and $\boldsymbol{r}_n$, respectively, relative to the origin, $\mathcal{O}$, are separated by $\Delta\boldsymbol{r}_{mn}$. Elements, $a$ and $b$, in stations, $m$ and $n$, are denoted by $a_m$ and $b_n$, located at $\boldsymbol{r}_{a_m}$ and $\boldsymbol{r}_{b_n}$, respectively. $\Delta\boldsymbol{r}_{a_m b_n}$ is the separation between these elements. The spatial sizes of the element, station, and the array in the aperture plane are denoted by $D_\textrm{e}$, $D_\textrm{s}$, and $D_\textrm{A}$, respectively. The solid angle, $\Omega_\textrm{e}\sim (\lambda/D_\textrm{e})^2$, corresponding to the element size determines the overall field of view in the sky plane accessible by the aperture array. Intra-station architectures produce station-sized virtual telescopes that simultaneously produce beams of solid angle, $\Omega_\textrm{s}\sim (\lambda/D_\textrm{s})^2$, in multiple directions, $\hat{\boldsymbol{s}}_k$, that can fill $\Omega_\textrm{e}$. Inter-station processing towards any of these beams produces virtual telescopes of size, $D_\textrm{A}$, that simultaneously produce beams of solid angle, $\Omega_\textrm{A}\sim (\lambda/D_\textrm{A})^2$, in multiple directions, $\boldsymbol{\sigma}_\ell$, relative to $\hat{\boldsymbol{s}}_k$, that can fill the solid angle of that beam. 
\label{fig:geometry-notation}}
\end{figure}

The number of stations in the array and the number of elements per station (indexed by $m$) are denoted by $N_\textrm{s}$ and $N_{\textrm{eps}_m}$, respectively. The number of elements in all stations are assumed to be equal, $N_{\textrm{eps}_m} = N_\textrm{eps}$, for convenience. Without loss of generality, this study can be extended to arrays with heterogeneous station densities and layouts such as LOFAR, but the idiosyncrasies of the array will have to be taken into account.

The specific aperture array examples considered are the \texttt{SKA-low}, the core of the \texttt{SKA-low} (\texttt{SKA-low-core}), the first phase of the concept of a long baseline extension to \texttt{SKA-low} called Low-frequency Australian Megametre Baseline Demonstrator Array (\texttt{LAMBDA-I}), the Compact All-Sky Phased Array \citep[\texttt{CASPA}\footnote{Presently, CASPA's stations are intended for localisation and astrometry, rather than for aperture synthesis.};][]{Luo+2024}, and the core of a lunar array \citep[\texttt{FarView};][]{Polidan+2024} which is hereafter referred to as \texttt{FarView-core}. \texttt{LAMBDA-I}, \texttt{CASPA}, and \texttt{FarView-core} are hypothetical interferometers that are still in concept phase.

@@

Note that, traditionally, cosmological applications such as probing the epoch of reionisation (EoR) using redshifted 21~cm have relied on storing the visibilities and processing them offline. In such cases, although imaging costs like gridding, Discrete Fourier Transform (DFT), or FFT will not apply, the correlations must still be obtained in real-time. With large arrays, the traditional correlator approach may be too expensive. For example, the Packed Ultra-wideband Mapping Array \citep[PUMA;][]{PUMA+2019}, a planned radio telescope for cosmology and transients with $\sim 5000-32000$ elements, is considering an FFT correlator architecture, a specific adaptation of the EPIC architecture for redundant arrays. This study, with a breakdown of costs for different operations, is also relevant for such cosmological arrays.

@@

The primary metric for computational cost used in this study is the number of real floating point operations (FLOP) and the density of such operations in the discovery phase space volume whose dimensions span time, frequency, polarisation, and location on the sky. The computational cost will be estimated per independent interval of time (in FLOP per second or FLOPS) for all possible polarisation states per frequency channel ($\delta\nu$) per independent angular resolution element ($\delta\Omega$). In this paper, a discovery phase space volume element comprising of a single frequency channel at a single two-dimensional angular resolution element (independent pixel) will be referred to as a \textit{voxel} ($\delta\nu\,\delta\Omega$).

@@

Two broad classes of two-stage fast imaging architectures are considered. In both classes, the first stage is intra-station interferometry of the element electric fields using the options of voltage beamforming (BF), E-field Parallel Imaging Correlator (EPIC), beamforming of correlations (XBF), or correlation followed by FFT (XFFT). The second stage involves inter-station combination of the data products from the first stage. The main difference between the two classes is in whether the second stage uses the first stage products coherently or incoherently. Although aperture arrays consisting of only three hierarchical scales are considered here, the analysis can be extended without loss of generality to arrays of even larger levels of spatial hierarchy. A schematic of these hybrid architectures, their computational cost components, and their extension methodology to multiple stages are illustrated in Figure~\ref{fig:hybrid-architectures-cost-schematic}.

\begin{figure}
\includegraphics[width=0.95\linewidth]
  % {figures/flops_calc_abridged_v2.pdf}
{fig_02.pdf}
\caption{A hybrid, two-stage architecture for hierarchical aperture arrays showing the methodology for extension to multiple stages. Black arrows indicate pathways where the signal consists of electric fields with phase coherence. Gray arrows indicate pathways of a ``squared'' or a correlated signal with no absolute but only relative phase information. At the top are intra-station architectures, namely, BF, correlator, and EPIC. The top box denotes intra-station imaging options. The bottom box denotes inter-station imaging architectures that process coherent outputs obtained from multiple stations deploying intra-station BF and/or EPIC architectures. The electric fields phase coherently synthesised by BF or EPIC at any stage can form the input for the next stage, thereby allowing the extension of the hierarchy.
\label{fig:hybrid-architectures-cost-schematic}}
\end{figure}

@@

In the discussions and equations that follow, a real-time calibration is assumed to have been performed on the input voltages. By assuming the system is reasonably stable, a much less frequent ($\ll 1/t_\textrm{acc}$) calibration will be sufficient, which therefore will not significantly affect the total cost budget \citep{Beardsley+2017,Gorthi+2021}. Although in principle an accurate calibration is possible in real time, applications such as EoR that require high-accuracy calibration have, for practical reasons, relied on offline calibration.

@@

% The intra-station architectures comprising the first stage form the basis for both approaches.
This first stage essentially synthesises station-sized apertures with simultaneous, multiple virtual pointings. These station-sized synthesised apertures from the first stage will act as building blocks for the second stage of inter-station processing. The element, $a$, in station, $m$ is denoted by $a_m$. The element and station locations are denoted by $\boldsymbol{r}_{a_m}$ and $\boldsymbol{r}_m$, respectively. Similarly, $\Delta\boldsymbol{r}_{a_m b_n}$ is the separation between an elements, $a$ and $b$, in stations, $m$ and $n$, respectively. $\Delta\boldsymbol{r}_{mn}$ denotes the station separation. This notation is illustrated in Figure~\ref{fig:geometry-notation}. A summary of the intra-station costs are provided in Table~\ref{tab:intra-inter-station-coherent-imaging}.

@@

Let $\widetilde{E}_{a_m}^{p}(\nu)$ represent the calibrated and potentially noise-weighted electric field at frequency, $\nu$, in polarisation state, $p$, measured by a station element, $a_m$, in station, $m$. Because each spectral channel is assumed to be processed independently, specifying the frequency can be dropped for convenience to write $\widetilde{E}_{a_m}^{p}(\nu)\equiv \widetilde{E}_{a_m}^{p}$. Then, such electric fields from individual station elements
% The calibrated electric fields, $\widetilde{E}_{a_m}^p$, recorded by individual elements, $a_m$, within a station, $m$, and polarisation state, $p$,
can be phase-coherently superposed to obtain the electric field in polarisation state, $\alpha$, towards any desired direction as
\begin{align}
    \widetilde{\mathcal{E}}_m^\alpha(\hat{\boldsymbol{s}}_k) &= \sum_{a_m} \sum_p  \widetilde{\mathcal{W}}_{a_m}^{\alpha p *}(\hat{\boldsymbol{s}}_k) \, \widetilde{E}_{a_m}^p \, e^{i\frac{2\pi}{\lambda} \hat{\boldsymbol{s}}_k\cdot\boldsymbol{r}_{a_m}} \, , \label{eqn:intra-station-pol-hol-img-expl}
\end{align}
where, $\hat{\boldsymbol{s}}_k$ denotes the direction of the superposed beam, $k$,
% $\widetilde{E}_{a_m}^{p}$ represents the calibrated and potentially noise-weighted electric field measured by the station element, $a_m$,
and $\widetilde{\mathcal{W}}_{a_m}^{{\alpha p}^*}(\hat{\boldsymbol{s}}_k)$ denotes a complex-valued directional weighting\footnote{One can choose $\widetilde{\mathcal{W}}_{a_m}^{\alpha p}(\hat{\boldsymbol{s}}_k)=\mathcal{W}_{a_m}^{p\alpha}(\hat{\boldsymbol{s}}_k)$, the directional electric field sensitivity, if the signal-to-noise ratio is to be optimised. But it can also be generically chosen depending on the property desired of the estimator \cite[][]{Morales2011}.} with the superscript indices $\alpha, p$, representing the contribution of polarisation state, $p$ in the station element, $a_m$, to polarisation state, $\alpha$, towards direction, $\hat{\boldsymbol{s}}_k$. The * denotes complex conjugate. Equation~(\ref{eqn:intra-station-pol-hol-img-expl}) resembles a DFT of the calibrated Electric field measurement after applying a complex-valued weighting.

The polarised intensity in the beamformed pixel is then obtained by
\begin{align}
    \widetilde{\mathcal{I}}^{\alpha\beta}_m(\hat{\boldsymbol{s}}_k) &= \left\langle \widetilde{\mathcal{E}}_m^\alpha(\hat{\boldsymbol{s}}_k) \,  \widetilde{\mathcal{E}}_m^{\beta *}(\hat{\boldsymbol{s}}_k) \right\rangle \, , \label{eqn:intra-station-opt-pol-img-outprod}
\end{align}
where, angular brackets denote a temporal averaging across an interval of $t_\textrm{acc}$. $\widetilde{\mathcal{I}}^{\alpha\beta}_m(\hat{\boldsymbol{s}}_k)$ results from the outer product of indices $\alpha$ and $\beta$. Superscript, $\alpha\beta$, thus denotes the four pairwise combinations of polarisation states of the intensity. Even though it is an outer product over the polarisation indices, it will be referred to as a ``squaring'' operation, hereafter, for convenience because of what it reduces to if $\alpha=\beta$.

@@

Figure~\ref{fig:1D-incoherent-compcost-BF-LAMBDA-I} shows a breakdown of the computational cost per station per voxel for voltage beamforming as a function of the different station parameters. In each panel, all parameters except the one on the $x$-axis are kept fixed at the characteristic values of the \texttt{LAMBDA-I}, \texttt{SKA-low}, and \texttt{SKA-low-core} stations as highlighted in cyan. The total cost is dominated by the DFT beamforming in Equation~(\ref{eqn:intra-station-pol-hol-img-expl}) denoted by black dashed lines. The DFT beamforming cost per voxel scales linearly with the number of elements per station. The costs of squaring and temporal averaging in Equation~(\ref{eqn:intra-station-opt-pol-img-outprod}) are subdominant. Because the squaring and temporal averaging are performed on a per-pixel basis and not dependent on $N_\textrm{eps}$, they remain constant across $N_\textrm{eps}$. And because the number of independent pixels covering the field of view scales as $n_\textrm{bs} \simeq (D_\textrm{s}/D_\textrm{e})^2$, the cost per independent pixel remains constant across $D_\textrm{s}$ and $D_\textrm{e}$. All these operations have to be performed at every time step, $\delta t$, and hence remain constant regardless of $t_\textrm{acc}$. Although spatial averaging across multiple stations (orange dot-dashed line) is not a station-level operation, it is shown to emphasise that it constitutes only a negligible cost relative to the other components, noting that it needs to be only performed on a slower cadence of $t_\textrm{acc}$ on a per-pixel basis, and thus scales inversely with $t_\textrm{acc}$.
